Large-scale spatiotemporal variation in vital rates of an alpine bird

### 2023-12-18 Manuscript_Nater_etal.qmd
Updated text and references (.bib file)

-   Setup of Hønsefugltakseringen monitoring programme (volunteer training, line transect sampling, spatio-temporal coverage, etc.)
-   Data & metadata recording via app
-   Quality control of data
-   Open access publishing of data (GBIF, LivingNorway)
-   Pre-modelling data wrangling

Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/). Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).

@@

-   Description of setup, data collection, temporal coverage, etc.
-   Pre-modelling data wrangling

\[End - ERLEND drafts\]

>>>>

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect willow ptarmigan observations along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy). When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age- and sex of the observed birds. An observation typically include 1 - 12 birds (mean = ?), representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Before 2019, data were collected on a dedicated field form, and entered manually in a web portal. After field data has been collected, several steps of quality control takes place before, both by local stakeholders and personnel from NINA. Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). This data set contain both meta data about the transect surveys (survey date, line transect length and location, study area ID and other information) as well as data pertaining to the observations (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). For this study, we used only data from public land that is freely available on GBIF; thus the workflow we describe should be fully reproducible by others. Formally, the data is publsihed as three distinct data sets, based on the three main public land owners (Statskog, FeFo and Fjellstyrene, respectively).

The program is not designed as a national monitoring, but rather a collection of local and regional survey programs. All involved survey areas use a common field protocol and data collection model. In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design is provided. However, both a local scale and in particular at a regional scale, the design is unbalanced due to the fact that participation in voluntary by stakeholders. In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway. In total, we used data from Xx different survey areas, across xx counties and xx municipalities.

@@

In our model, we integrated data from an ongoing field study of marked birds in Lierne municipality in Central Norway. These data were used to inform the survival component of the model (se below). The data is based on a sample of radio marked birds. The birds were captured in winter (late February or early March each year in the study period), and fitted with VHF collars. The marked birds were monitored on a regular basis until they either i) died, ii) battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored at least monthly. However, during the breeding and chick-rearing season (May to July) the birds were monitored more often, and during december and January we obtained few observations due to challenging field work conditions. A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes. We used data from the years 2015 - 2021, and the total sample size across these years was xx birds. The study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
